Ensembler: Enabling high-throughput molecular simulations at the superfamily scale

### 2015-03-22 manuscript/ms.tex
Reworking some figure captions.

As a first application of {\bf Ensembler}, we have built models for all 90 human tyrosine kinase (TK) domains listed in UniProt.
{\color{red}[JDC: Is there a complete list of these somewhere?  Maybe reference supplementary data?]}
TKs (and protein kinases in general) play important roles in many cellular processes and are involved in a number of types of cancer.
{\color{red}[JDC: CITE]}
For example, mutations of Src are associated with colon, breast, and prostate cancer {\color{red}[CITE: Src cancer involvement]}, while a translocation between the TK Abl1 and the pseudokinase Bcr is closely associated with chronic myelogenous leukemia {\color{red}[CITE: Abl1 cancer involvement]}.
Protein kinase domains are thought to have multiple accessible metastable conformation states, with a single active conformation, and much effort is directed at developing kinase inhibitor drugs which bind to and stabilize inactive conformations [CITE: Lee and Craik Science 2009].
{\color{red}[JDC: Lee and Craik do not discuss kinases, I don't believe; you'll have to find an accurate reference on kinase conformations.]}
Kinases are thus a particularly interesting subject for study with MSM methods [CITE: recent kinase MSM papers], and this approach stands to benefit greatly from the ability to exploit the full body of available genomic and structural data within the kinase family, e.g.~by generating large numbers of starting configurations to be used in highly parallel MD simulation.

@@

Crystallographic structures of kinase catalytic domains generally contain a significant number of missing residues (median 11, standard deviation 13, max 102) due to the high mobility of several loops (Fig.~\ref{figure:loopmodel-nmissing-residues}, top), with a number of these missing spans being significant in length (Fig.~\ref{figure:loopmodel-nmissing-residues}, bottom).
To reduce the reliance on the Modeller rapid model construction stage to reconstruct very long unresolved loops, unresolved template residues were first remodeled using the {\tt loopmodel} subcommand.
Out of \num{3666} templates with one or more missing residues, \num{3134} were successfully remodeled by the Rosetta loop modeling stage (with success defined simply as program termination with out error); most remodeling failures were attributable to unsatisfiable spatial constraints imposed by the original template structure.
There was some correlation between remodeling failures and the number of missing residues (Fig.~\ref{figure:loopmodel-nmissing-residues}, top); templates for which remodeling failed had a median of 20 missing residues, compared to a median of 14 missing residues for templates for which remodeling was successful.

@@

The distribution of RMSDs of the final models (relative to the highest sequence identity model for a given target) is shown in Fig.~\ref{figure:rmsd-distribution-by-sequence-identity}.
The distributions are stratified based on the sequence identity between target and template, indicating that higher sequence identity templates result in models with lower RMSDs.
The sequence identity stratifications were selected based on the sequence identity distribution plotted in Fig.~\ref{figure:sequence-identity-distribution}, which suggests an intuitive division into three categories, with \num{307753} models in the 0-35\% sequence identity range, \num{69922} models in the 35-55\% range, and \num{4893} models in the 55-100\% range.

To provide a more complete evaluation of the models generated, we have analyzed two example TKs (\emph{Src} and \emph{Abl1}) in detail.
Due to their importance in cancer, as outlined above, these kinases have been the subject of numerous studies, encompassing many different methodologies.
In terms of structural data, a large number of crystal structures have been solved (with or without ligands such as nucleotide substrate or inhibitor drugs), showing the kinases in a number of different conformations.
These two kinases are thus also interesting targets for MSM studies, with one recent study focusing on modeling the states which constitute the activation pathway of Src {\color{blue}[CITE:Shukla Pande Nat Commun 2014]}.

Fig.~\ref{figure:superposition} shows a superposition of a set of representative models of \emph{Src} and \emph{Abl1}.
Models were first stratified into three ranges, based on the structure of the sequence identity distribution (Fig.~\ref{figure:sequence-identity-distribution}), then subjected to $k$-medoids clustering to pick three representative models from each sequence identity range.
{\color{red}[JDC: Explain how $k$-medoids clustering was done either here or in figure caption.]}
Each model is colored and given a transparency based on the sequence identity between the target and template sequence.
The figure gives an idea of the variance present in the generated models.
High sequence identity models (in opaque blue) tend to be quite structurally similar, with some variation in loops or changes in domain orientation.
The Abl1 renderings indicate one high sequence identity model with a long unstructured region at one of the termini, which was unresolved in the original template structure.
While such models are not necessarily incorrect or undesirable, it is important to be aware of the effects they may have on production simulations performed under periodic boundary conditions, as long unstructured termini can be prone to interact with a protein's periodic image.
Lower sequence identity models (in transparent white or red) indicate much greater variation in all parts of the structure.
We believe the mix of high and low sequence identity models to be particularly useful for methods such as MSM building, which require thorough sampling of the conformational landscape.
The high sequence identity models could be considered to be the most likely to accurately represent true metastable states.
Conversely, the lower sequence identity models could be expected to help push a simulation into regions of conformation space which might take intractably long to reach if starting a single metastable conformation.

To evaluate the models of \emph{Src} and \emph{Abl1} in the context of the published structural biology literature, we have focused on two residue pair distances thought to be important for the regulation of protein kinase domains.
We use the residue numbering schemes for chicken Src (which is commonly used in the literature even in reference to human Src)[CITE: 2SRC, 1Y57] and human Abl1 isoform A[CITE: 2F4J, 2HYY, 2G1T] respectively; the exact numbering schemes are provided in Supporting Information S\ref{si:residue-numbering-schemes}.
Fig.~\ref{figure:src-ref-structures} shows two structures of \emph{Src} believed to represent inactive (PDB code: 2SRC) {\color{red}[CITE: 2SRC]} and active (PDB code: 1Y57) {\color{red}[CITE: 1Y57]} states.
One notable feature which distinguishes the two structures is the transfer of an electrostatic interaction of E310 from R409 (in the inactive state) to K295 (in the active state), brought about by a rotation of the $\alpha$C-helix.
These three residues are also well conserved [CITE Kannan Neuwald JMB 2005], and a number of experimental and simulation studies have suggested that this electrostatic switching process plays a role in a regulatory mechanism shared across the protein kinase family [CITE Foda Shan Seeliger Src Nat Commun 2015; Shukla Pande Nat Commun 2014; Ozkirimli Post Prot Sci 2008].
As such, we have projected the {\bf Ensembler} models for \emph{Src} and \emph{Abl1} onto a space consisting of the distances between these two residue pairs (Fig.~\ref{figure:pair-distances}).
The models show strong coverage of regions in which either of the electrostatic interactions is formed, as well as a wide range of regions inbetween.
We thus expect that such a set of models, if used as starting configurations for highly parallel MD simulation, could greatly aid in sampling of the activation process.

@@

    \caption{{\bf Sequence identity distribution for human TK models.}
    Distribution of sequence identities for all \num{373513} models generated for the human tyrosine kinases.
    Sequence identities are calculated from pairwise target-template alignments.
    The cumulative distribution function is shown by the dashed line.
    The plotted distributions have been smoothed using kernel density estimation.
    % {\color{red}[JDC: Font size too small.]}
    }
  \label{figure:sequence-identity-distribution}
\end{figure}

\begin{figure}[tb]
    \includegraphics[width=1.0\columnwidth]{rmsddist/rmsddist2.pdf}

    \caption{{\bf RMSD distribution by sequence identity.}
    RMSD distributions for all \num{373513} human TK models, divided into three sequence identity ranges.
    For a given target, model RMSDs are calculated relative to the highest sequence identity model for that target.
    The plotted distributions have been smoothed using kernel density estimation.
    % {\color{red}[JDC: Font size too small.]}
  }
  \label{figure:rmsd-distribution-by-sequence-identity}
\end{figure}

@@

The code for {\bf Ensembler} is hosted on the collaborative open source software development platform GitHub,\\
\url{http://github.com/choderalab/ensembler} \\
The latest release of {\bf Ensembler} can be installed via the {\tt conda} package manager for Python~\cite{conda}:
\shellcmd{conda install -c https://conda.binstar.org/omnia ensembler}
This will install all dependencies except for Modeller and Rosetta, which are not available through the {\tt conda} package manager, and thus must be installed separately by the user.
The latest source can be downloaded from the GitHub repository, which also contains up-to-date instructions for building and installing the code.
{\color{red}[JDC: Where does documentation reside?  And example inputs for generating the results in this paper?]}

>>>>

As a first application of {\bf Ensembler}, we have built models for all 90 human tyrosine kinase (TK) domains listed in UniProt.
{\color{red}[JDC: What query was used to yield this set?]}
{\color{red}[JDC: Is there a complete list of these somewhere?  Maybe reference supplementary data?]}
TKs (and protein kinases in general) play important roles in many cellular processes and are involved in a number of types of cancer.
{\color{red}[JDC: CITE]}
For example, mutations of Src are associated with colon, breast, and prostate cancer {\color{red}[CITE: Src cancer involvement]}, while a translocation between the TK Abl1 and the pseudokinase Bcr is closely associated with chronic myelogenous leukemia {\color{red}[CITE: Abl1 cancer involvement]}.
Protein kinase domains are thought to have multiple accessible metastable conformation states, with a single active conformation, and much effort is directed at developing kinase inhibitor drugs which bind to and stabilize inactive conformations [CITE: Lee and Craik Science 2009].
{\color{red}[JDC: Lee and Craik do not discuss kinases, I don't believe; you'll have to find an accurate reference on kinase conformations.]}
Kinases are thus a particularly interesting subject for study with MSM methods [CITE: recent kinase MSM papers], and this approach stands to benefit greatly from the ability to exploit the full body of available genomic and structural data within the kinase family, e.g.~by generating large numbers of starting configurations to be used in highly parallel MD simulation.

@@

Crystallographic structures of kinase catalytic domains generally contain a significant number of missing residues (median 11, standard deviation 13, max 102) due to the high mobility of several loops (Fig.~\ref{figure:loopmodel-nmissing-residues}, top), with a number of these missing spans being significant in length (Fig.~\ref{figure:loopmodel-nmissing-residues}, bottom).
{\color{red}[JDC: Can you add statistics (median, stddev, max) for loop lengths too?]}
To reduce the reliance on the Modeller rapid model construction stage to reconstruct very long unresolved loops, unresolved template residues were first remodeled using the {\tt loopmodel} subcommand.
Out of \num{3666} templates with one or more missing residues, \num{3134} were successfully remodeled by the Rosetta loop modeling stage (with success defined simply as program termination with out error); most remodeling failures were attributable to unsatisfiable spatial constraints imposed by the original template structure.
There was some correlation between remodeling failures and the number of missing residues (Fig.~\ref{figure:loopmodel-nmissing-residues}, top); templates for which remodeling failed had a median of 20 missing residues, compared to a median of 14 missing residues for templates for which remodeling was successful.

@@

\begin{figure}[tbp]
    \includegraphics[width=1.0\columnwidth]{rmsddist/rmsddist2.pdf}

    \caption{{\bf Distribution of RMSDs to all tyrosine kinase catalytic domain models relative to model derived from highest sequence identity template.}    
    Distributions are averaged over all 90 tyrosine kinase catalytic domains.
    {\color{red}[JDC: Is this accurate?]}
    To better illustrate how conformational similarity depends on sequence identity, these were separated into three sequence identity classes: high identity (55--100\%), moderate identity (35--55\%), and remote identity (0--35\%).
    The plotted distributions have been smoothed using kernel density estimation.
    {\color{red}[JDC: Are these computed for the templates or the resulting models?]}
    {\color{red}[JDC: Can we also show the overall distribution without stratification (e.g.~in grey in a separate panel)?]}
  }
  \label{figure:rmsd-distribution-by-sequence-identity}
\end{figure}

To evaluate the diversity of conformations captured for each target sequence, we first computed the RMSD distributions for all models for each target (relative to the model derived from the highest-identity template) are shown in Fig.~\ref{figure:rmsd-distribution-by-sequence-identity}.
To better understand the influence of sequence identity on the conformational similarities of resulting models, the sequence identities were stratified based on the sequence identity distribution plotted in Fig.~\ref{figure:sequence-identity-distribution}, which suggests an intuitive division into three categories, with \num{307753} models in the 0--35\% sequence identity range, \num{69922} models in the 35--55\% range, and \num{4893} models in the 55--100\% range. 
It is clear that higher sequence identity templates result in models with lower RMSDs, while templates with remote sequence identities result in larger RMSDs on average.

To provide a more complete evaluation of the models generated, we have analyzed two example TKs (\emph{Src} and \emph{Abl1}) in detail.
Due to their importance in cancer, as outlined above, these kinases have been the subject of numerous studies, encompassing many different methodologies.
In terms of structural data, a large number of crystal structures have been solved (with or without ligands such as nucleotide substrate or inhibitor drugs), showing the kinases in a number of different conformations.
These two kinases are thus also interesting targets for MSM studies, with one recent study focusing on modeling the states which constitute the activation pathway of Src~\cite{shukla-pande:nature-commun:2014:src-activation-msm}.
Fig.~\ref{figure:superposition} shows a superposition of a set of representative models of \emph{Src} and \emph{Abl1}.
Models were first stratified into three ranges, based on the structure of the sequence identity distribution (Fig.~\ref{figure:sequence-identity-distribution}), then subjected to $k$-medoids clustering to pick three representative models from each sequence identity range.
{\color{red}[JDC: Explain how $k$-medoids clustering was done either here or in figure caption.]}
Each model is colored and given a transparency based on the sequence identity between the target and template sequence.
The figure gives an idea of the variance present in the generated models.
High sequence identity models (in opaque blue) tend to be quite structurally similar, with some variation in loops or changes in domain orientation.

The Abl1 renderings indicate one high sequence identity model with a long unstructured region at one of the termini, which was unresolved in the original template structure.
While such models are not necessarily incorrect or undesirable, it is important to be aware of the effects they may have on production simulations performed under periodic boundary conditions, as long unstructured termini can be prone to interact with a protein's periodic image.
Lower sequence identity models (in transparent white or red) indicate much greater variation in all parts of the structure.
We believe the mix of high and low sequence identity models to be particularly useful for methods such as MSM building, which require thorough sampling of the conformational landscape.
The high sequence identity models could be considered to be the most likely to accurately represent true metastable states.
Conversely, the lower sequence identity models could be expected to help push a simulation into regions of conformation space which might take intractably long to reach if starting a single metastable conformation.

To evaluate the models of \emph{Src} and \emph{Abl1} in the context of the published structural biology literature, we have focused on two residue pair distances thought to be important for the regulation of protein kinase domains.
We use the residue numbering schemes for chicken Src (which is commonly used in the literature even in reference to human Src)[CITE: 2SRC, 1Y57] and human Abl1 isoform A[CITE: 2F4J, 2HYY, 2G1T] respectively; the exact numbering schemes are provided in Supporting Information S\ref{si:residue-numbering-schemes}.
Fig.~\ref{figure:src-ref-structures} shows two structures of \emph{Src} believed to represent inactive (PDB code: 2SRC) {\color{red}[CITE: 2SRC]} and active (PDB code: 1Y57) {\color{red}[CITE: 1Y57]} states.
One notable feature which distinguishes the two structures is the transfer of an electrostatic interaction of E310 from R409 (in the inactive state) to K295 (in the active state), brought about by a rotation of the $\alpha$C-helix.
These three residues are also well conserved [CITE Kannan Neuwald JMB 2005], and a number of experimental and simulation studies have suggested that this electrostatic switching process plays a role in a regulatory mechanism shared across the protein kinase family~\cite{shukla-pande:nature-commun:2014:src-activation-msm} [CITE Foda Shan Seeliger Src Nat Commun 2015; Ozkirimli Post Prot Sci 2008].
As such, we have projected the {\bf Ensembler} models for \emph{Src} and \emph{Abl1} onto a space consisting of the distances between these two residue pairs (Fig.~\ref{figure:pair-distances}).
The models show strong coverage of regions in which either of the electrostatic interactions is formed, as well as a wide range of regions inbetween.
We thus expect that such a set of models, if used as starting configurations for highly parallel MD simulation, could greatly aid in sampling of the activation process.

@@

    \caption{{\bf Template-target sequence identity distribution for human tyrosine kinase catalytic domains.}
    Sequence identities are calculated from all pairwise target-template alignments, where targets are human kinase catalytic domain sequences and templates are all kinase catalytic domains from any organism with structures in the PDB, as described in the text.
    A kernel density estimate of the target-template sequence identity probability density function is shown as a solid line with shaded region, while the corresponding cumulative distribution function is shown as a dashed line.
    }
  \label{figure:sequence-identity-distribution}
\end{figure}

@@

The code for {\bf Ensembler} is hosted on the collaborative open source software development platform GitHub,\\
\url{http://github.com/choderalab/ensembler} \\
The latest release of {\bf Ensembler} can be installed via the {\tt conda} package manager for Python [\url{http://conda.pydata.org}]:
%\shellcmd{conda install -c https://conda.binstar.org/omnia ensembler}
\shellcmd{conda config --add channels https://conda.binstar.org/omnia}
\shellcmd{conda ensembler}
This will install all dependencies except for Modeller and Rosetta, which are not available through the {\tt conda} package manager, and thus must be installed separately by the user.
The latest source can be downloaded from the GitHub repository, which also contains up-to-date instructions for building and installing the code.
{\color{red}[JDC: Where does documentation reside?  And example inputs for generating the results in this paper?]}
